---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
Activity <- read.csv('activity.csv')
```


## What is mean total number of steps taken per day?
```{r}
TotalStepsPerDay <- aggregate(Activity$steps, list(Activity$date), sum)
hist(TotalStepsPerDay$x, 
     xlab = "Steps per Day", 
     ylab = "Number of Days", 
     main = "Frequency of Total Steps per Day", 
     breaks = "Sturges")
```

### Mean total steps per day
```{r}
# mean total steps per day
mean(TotalStepsPerDay$x, na.rm = T)
```

### Median total steps per day
```{r}
# median total steps per day
median(TotalStepsPerDay$x, na.rm = T)
```


## What is the average daily activity pattern?
```{r}
AvgStepsPerInterval <- aggregate(Activity$steps, 
                             list(Activity$interval), 
                             mean, 
                             na.rm = T, 
                             na.action = na.omit)
plot(AvgStepsPerInterval, 
     type = "l", 
     xlab = "5-minute Interval Code", 
     ylab = "Avg Num of Steps",
     main = "Average Steps for Each 5-minute Interval")

# Interval with most steps per day on average
AvgStepsPerInterval[AvgStepsPerInterval$x == max(AvgStepsPerInterval$x), 1]
```


## Imputing missing values
```{r}
Activity.NA.bool <- is.na(Activity$steps)
Activity.NA <- Activity[Activity.NA.bool,]
```

### Number of intervals with no data
```{r}
# Number of intervals with no data
length(Activity.NA[,1])
```

### Imputation strategy
Missing values for steps in a given interval are imputed using the mean number
of steps for that interval averaged across all days
```{r}
ImputedActivity <- Activity

# iterate through each row of dataset
for(i in 1:nrow(ImputedActivity)) 
{
    # check to see if row contains NA
    if (is.na(ImputedActivity[i, 1])) 
    {
        # find the corresponding interval
        interval = ImputedActivity[i, 3] 
        # find the average steps for this interval
        avgSteps = AvgStepsPerInterval[AvgStepsPerInterval$Group.1 == interval,2]
        # update the NA value with the average steps
        ImputedActivity[i,1] <- avgSteps
    }
}

ImputedTotalStepsPerDay <- aggregate(ImputedActivity$steps, list(ImputedActivity$date), sum)
hist(ImputedTotalStepsPerDay$x, 
     xlab = "Steps per Day", 
     ylab = "Number of Days", 
     main = "Frequency of Total Steps per Day with NA Values Imputed", 
     breaks = "Sturges")
# mean total steps per day
mean(ImputedTotalStepsPerDay$x, na.rm = T)
# median total steps per day
median(ImputedTotalStepsPerDay$x, na.rm = T)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
ImputedActivityFactored <- ImputedActivity
# factor dates to either "weekday" or "weekend"
ImputedActivityFactored$TypeOfDay <- apply(ImputedActivity, 1, function(row) {
    day = weekdays(as.POSIXct(row[2]))
    if (day %in% c("Saturday", "Sunday")) {
        as.factor("Weekend")
    }
    else {
        as.factor("Weekday")
    }
})

# aggregate average steps by interval code and weekday/weekend factor
ImputedActivityFactoredAvgSteps <- aggregate(
    ImputedActivityFactored$steps, 
    list(Interval = ImputedActivityFactored$interval, TypeOfDay = ImputedActivityFactored$TypeOfDay), 
    mean, 
    na.rm = T, 
    na.action = na.omit)

library(ggplot2)

# plot avg number of steps against interval code, using panels for weekday/weekend
ggplot(ImputedActivityFactoredAvgSteps, aes(Interval, x)) + 
    facet_grid(TypeOfDay ~ .) + 
    geom_line() +
    xlab("5-min Interval Code") + 
    ylab("Number of Steps") +
    ggtitle("Average Number of Steps at Each Interval for Weekdays and Weekends")
```

There are differences in the activity pattern between weekdays and weekends. 

During weekdays there is greater activity from 5AM till about 9AM as compared 
to weekends. 

There is more activity during the middle portion of the day in the weekends 
compared to the weekdays.

The activity continues later in the day during weekends than during weekdays.